Gaussian Process regression for astronomical time-series

### 2022-08-29 src/tex/ms.tex
[showyourwork] overleaf sync

@@

Perhaps the earliest use of GPR in the refereed astronomical literature that will be familiar to a modern reader was published by \citet{prh92a} in the context of quasar variability, and for a long time this remained its main application domain in astronomy. GPR then gradually appeared in other areas, starting with photometric redshift estimation \citep{2006ApJ...647..102W}, then exoplanet transit observations \citep{2009ApJ...704...51C,2012MNRAS.419.2683G} and radial velocity planet searches \citep{2012MNRAS.419.3147A,2014MNRAS.443.2517H}. Nonetheless, GPR remained relatively niche and few astronomers had heard of it until a few years ago. To illustrate this point, we searched on the NASA Astrophysics Data System (ADS) for articles published in refereeed astronomy and astrophysics journals with the words ``Gaussian Process'' in full text of the article \autoref{fig:literature}.
After an increase in popularity throughout the 1990s, the use of GPs in astrophysics remained fairly constant around $\sim20$ publications per year until 2010.
Since 2010, the popularity of GPs has grown significantly, and in 2021, more than 500 refereed papers referencing GPs were published in astrophysics literature.

@@

In this second example, we demonstrate another common application of GPR in astrophysics, as a flexible model for nuisances and correlated noise where we want to correctly capture uncertainty introduced by this noise model into our constraints on the parameters of interest.
To this end, we simulate the light curve of a transiting exoplanet and aim to infer the physical parameters of the system, taking correlated noise into account.
The correlated noise in transit light curves is typically caused by the variability of the host star and by instrumental effects like focus or pointing changes.
It has been demonstrated that neglecting to account for these variations can cause significant errors when inferring the properties of the planet \dan{add references}.
This correlated noise can often be well-modeled by a GP, and the use of GPs for transit modeling has been a fruitful area of research in the astrophysics literature \dan{add references}.

For this example, the transit was simulated with known physical properties such as the planet-to-star radius ratio, the impact parameter, and quadratic limb darkening parameters \dan{cite Agol for light curve computation}.
We then add correlated and white noise by sampling from a GP model with a Mat\'ern-3/2 covariance function as defined in \autoref{sec:kernels}, with a known amplitude, timescale, and white noise amplitude.
These simulated data are shown in the left panel of \autoref{fig:transit}

@@

The parameters $\meanparams$ and $\hyperparams$ of the mean and covariance function are known as the \textit{hyper-parameters} of the GP. Strictly speaking, the \textit{parameters} of the GP are the elements of the mean vector and covariance matrix. However, for a given set of inputs and choice of mean and covariance functions, these numerous individual parameters are entirely determined by the the hyper-parameters, so the user interacts with the latter in much the same way as they would with the parameters of a more classical model.

Although is is possible to construct and use stochastic process models based on other distributions, GPs are by far the most popular, for two main reasons. The first is Central Limit theorem: it implies that the assumption of Gaussianity is often at least approximately correct. The second is that Gaussian distributions obey simple mathematical identities for marginalisation and conditioning, that enable inference with GPs to be performed analytically, as we will see later.

@@

Consider $N$ observations of a variable $\mathbf{y} = \{ y_i\}_{i=1, \ldots, N}$, taken at times $\mathbf{t} = \{ t_i\}$, with associated measurement uncertainties $\boldsymbol{\sigma}=\{\sigma_i\}$. We wish to compare these to a model function $m(t,\meanparams)$ controlled by parameters $\meanparams=\{\meanparams_j\}_{j=1,\ldots,M}$. In most astrophysical applications, we are interested in estimating some or all of those parameters. In least-squares regression, we minimize the quantity
\begin{equation}
  \chi^2 \equiv\sum_{i=1}^N (y_i-m_i)^2/\sigma^2_i,
\end{equation}
where $m_i \equiv m(x_i,\meanparams)$, with respect to $\meanparams$. Where does this come from?

@@

\subsubsection{Link to Gaussian Process Regression}\label{sec:gp-lsq-link}
\dan{I think we should explicitly re-write this as $\mathcal{L}(\meanparams,\,\hyperparams)$ and indicate that $K$ and $m$ depend on these parameters. We should really highlight that this is a drop-in replacement for the likelihood defined in the previous section.}
Let us now re-write the likelihood in matrix form:
\begin{equation}
\label{eq:gp_like}
  \mathcal{L} = 
  \frac{1}{\sqrt{|2\,\pi\,\mathbf{K}|}} \exp \left(-(\mathbf{y}-\mathbf{m})^{\mathrm{T}} \mathbf{K}^{-1}
  (\mathbf{y}-\mathbf{m}) \right),
\end{equation}
where $\mathbf{K}$ is a purely diagonal ($N$, $N$) matrix with elements $K_{ij} = \delta_{ij} \sigma_i^2$ ($\delta_{ij}$ being the discrete Kronecker delta function). This is, of course, the \emph{covariance matrix} of the model.

Now, instead of assuming that the covariance matrix takes this very specific form, let us allow a more flexible model for the covariance:
\begin{equation}
  K_{ij} = k(t_i,t_j,\hyperparams) + \delta_{ij} \sigma_i^2,
\end{equation}
where $k$ is a \emph{covariance function} or \emph{kernel}, controlled by parameters $\hyperparams$. The result is a GP, and its likelihood is still given by Equation~(\ref{eq:gp_like}). Depending on the choice of kernel and parameters, the covariance matrix can now have non-zero off-diagonal elements, allowing us to explicitly model correlated noise or stochastic signals in the data.
Note that the likelihood depends not only on the argument of the exponential, but also on the determinant of the covariance matrix, which therefore needs to be evaluated explicitly. This term acts as a built-in Occam's razor, automatically penalising more complex models.

@@

\subsubsection{Properties of the  predictive distribution}
A closer look at Equations~(\ref{eq:pred}) reveals some important properties.
First, a GP is a \textit{linear predictor}. The predictive mean for a specific time $t_\star$ can be written as a linear combination of the observations:
$f_\star = \mathbf{w}^{\mathrm{T}} \, \mathbf{y}$, where $\mathbf{w} = \mathbf{K}^{-1} \, \mathbf{k}_\star$ and $\mathbf{k}_\star=k(\mathbf{t},t_\star,\hyperparams)$. It can also be written as
a linear combination of covariance functions centred on the training points:
$f_\star = \mathbf{\alpha}^\mathrm{T} \,  \mathbf{k}_\star$ where $\mathbf{\alpha} = \mathbf{K}^{-1} \, \mathbf{y}$.
These linearity properties are very important to understand the behaviour of GPs. The shed light on the relationship between GPR and standard linear models with very large numbers of free parameters \citep[see e.g.][for a more detailed discussion]{2021PASP..133i3001H},
and lead to a number of powerful extensions that are beyond  scope of this review.

@@

In the previous section, we presented an overview of GP methods, and the key mathematical details. In this section, we will dive deeper into some of the practical decisions that arise when using GPs.
The two core elements of a GP model are the mean function $m(t;\,\meanparams)$, and the ``kernel'' or ``covariance'' function $k(t_i,\,t_j;\,\hyperparams)$.
Unlike how they are typical presented in the machine learning literature, GP models in astrophysics will often---but not always---include non-trivial mean functions.
For example, in the example from \autoref{sec:transit}, the mean function $m(t;\,\meanparams)$ is a physical transit model that is a function of the orbital parameters of the system, and includes a realistic limb-darkening model \citep{2002ApJ...580L.171M}.
However, in this review we won't discuss the mean function in detail, focusing instead primarily on the kernel function, since that is unique to GP modeling.
All this being said, in our experience, new users of GP models will often focus and worry more than necessary about the choice of kernel function for their problem.
As with any probabilistic modeling problem, there are several well-defined workflows for motivating, selecting, and validating your choice of kernel.
In this section we walk through this process in detail.

@@

The kernel function can be any positive semi-definite scalar function, but some will be more useful than others.
% A commonly (over-)used kernel function is the ``exponential-squared'' or ``radial basis function'' kernel
% \begin{equation}
%   k(t_i,\,t_j;\,\hyperparams) = \sigma^2\,\exp\left[-\frac{(t_i - t_j)^2}{2\,\tau^2}\right]
% \end{equation}
% where the ``hyperparameters'' $\hyperparams$ are $\sigma$ and $\tau$.
% In this case, the hyperparameters specify the typical amplitude $\sigma$ and timescale $\tau$ of the variance of the process.
% SOMEFIGURE shows how changing these parameters affect our GP model.
% There exist many other commonly used kernel functions, some of which are listed in SOMETABLE.
% This table is far from comprehensive, and importantly, the sum or product of two valid kernel functions is itself also a valid kernel, so it is possible (and common) to build extremely flexible and expressive kernel functions specific to your use case.
Given this large decision space, you may be wondering how to choose the right kernel for your specific problem.
If you're very lucky, you may be able to motivate your model using physics.
This can be approached from two directions.
On one hand, some commonly used kernel function have a specific physical interpretation---for example, the solution to a stochastic ordinary differential equation---that can be used to motivate their use.
On the other hand, if your physical model is specified by its \emph{power spectrum}, this can recast as a GP model with a specific kernel function.

@@

Some popular kernels are listed in \autoref{tab:kernels}, and some other choices are discussed in Chapter 4 of \citet{gpml}.
In any practical application, these kernel functions are not used on their own.
Instead, more expressive models are designed by combining these models.
In particular, any valid kernel functions can be added or multiplied to generate new valid functions.
For example, the squared exponential kernel function is generally defined as
\begin{equation}
  k(\dt;\,\hyperparams) = \amplitude^2\,\exp\left(-\frac{\dt^2}{2\,\lengthscale^2}\right)
\end{equation}
with $\hyperparams = \{\amplitude,\,\lengthscale\}$, which, in our notation, is actually the product of a constant kernel and a squared exponential kernel, as defined in \autoref{tab:kernels}.
Another example that is commonly used in the astrophysics literature \dan{ADD SOME CITATIONS} is the following ``quasi-periodic'' kernel
\begin{equation}
  k(\dt;\,\hyperparams) = \amplitude^2\,\exp\left(-\frac{\dt^2}{2\,{\lengthscale_1}^2} -\gamma\,\sin^2\left[\frac{\pi\,\dt}{\lengthscale_2}\right] \right)
\end{equation}
with $\hyperparams = \{\amplitude,\,\lengthscale_1,\,\lengthscale_2,\,\gamma\}$, which has a period of $\lengthscale_2$, and a decoherence timescale of $\lengthscale_1$.

@@

@@

\subsection{Stars and exoplanets}

@@

\subsubsection{Transit searches}

@@

\subsubsection{Transit and eclipse modelling}

>>>>

\suz{Should we mention we are using showyourwork?}

@@

Perhaps the earliest use of GPR in the refereed astronomical literature that will be familiar to a modern reader was published by \citet{prh92a} in the context of quasar variability, and for a long time this remained its main application domain in astronomy. GPR then gradually appeared in other areas, starting with photometric redshift estimation \citep{2006ApJ...647..102W}, then exoplanet transit observations \citep{2009ApJ...704...51C,2012MNRAS.419.2683G} and radial velocity planet searches \citep{2012MNRAS.419.3147A,2014MNRAS.443.2517H}. Nonetheless, GPR remained relatively niche and few astronomers had heard of it until a few years ago. To illustrate this point, we searched on the NASA Astrophysics Data System (ADS) for articles published in refereeed astronomy and astrophysics journals with the words ``Gaussian Process'' in full text of the article \autoref{fig:literature}.
After an increase in popularity throughout the 1990s, the use of GPs in astrophysics remained fairly constant around $\sim20$ publications per year until 2010.
Since 2010, the popularity of GPs has grown significantly, and in 2021, more than 500 refereed papers referencing GPs were published in the astrophysics literature.

@@

In this second example, we demonstrate another common application of GPR in astrophysics, as a flexible model for nuisances and correlated noise where we want to correctly capture uncertainty introduced by this noise model into our constraints on the parameters of interest.
To this end, we simulate the light curve of a transiting exoplanet and aim to infer the physical parameters of the system, taking correlated noise into account.
The correlated noise in transit light curves is typically caused by the variability of the host star and by instrumental effects like focus or pointing changes.
It has been demonstrated that neglecting to account for these variations can cause significant errors when inferring the properties of the planet \citep{2006MNRAS.373..231P,2007A&A...472L..13G}.
This correlated noise can often be well-modeled by a GP, and the use of GPs for transit modeling has been a fruitful area of research in the astrophysics literature (see Section~\ref{sec:transit_fit} for a more detailed discussion).

For this example, the transit was simulated with known physical properties such as the planet-to-star radius ratio, the impact parameter, and quadratic limb darkening parameters \dan{cite Agol for light curve computation}.
We then add correlated and white noise by sampling from a GP model with a Mat\'ern-3/2 covariance function as defined in \autoref{sec:kernels}, with a known amplitude, timescale, and white noise amplitude.
These simulated data are shown in the left panel of \autoref{fig:transit}.

@@

The parameters $\meanparams$ and $\hyperparams$ of the mean and covariance function are known as the \textit{hyper-parameters} of the GP. Strictly speaking, the \textit{parameters} of the GP are the (infinitely many) unknown functions that share the specified mean vector and covariance matrix and could have given rise to the observations. However, these parameters are always marginalised over: we never explicitly deal with the individual functions, except when drawing samples for illustrative purposes (as we did in Figures~\ref{fig:quasar} and \ref{fig:transit}). GPs are therefore a type of \textit{Hierarchichal Bayesian hierarchical Model} (HBM).

Although is is possible to construct and use stochastic process models based on other distributions, GPs are by far the most popular, for two main reasons. The first is Central Limit theorem: it implies that the assumption of Gaussianity is often at least approximately correct. The second is that Gaussian distributions obey simple mathematical identities for marginalisation and conditioning, that inference with GPs (the process of marginalising over the individual functions) to be performed \textit{analytically}, with very simple linear algebra. It is this analytic marginalisation property that sets GPR apart from other forms of HBM. \suz{ask Dan to check these last two paragraphs.}

@@

Consider $N$ observations of a variable $\mathbf{y} = \{ y_i\}_{i=1, \ldots, N}$, taken at times $\mathbf{t} = \{ t_i\}$, with associated measurement uncertainties $\boldsymbol{\sigma}=\{\sigma_i\}$. We wish to compare these to a model function $m(t,\meanparams)$ controlled by parameters $\meanparams=\{\meanparams_j\}_{j=1,\ldots,M}$. In most astrophysical applications, we are interested in estimating some or all of those parameters. In least-squares regression, we minimize the quantity
\begin{equation}
  \chi^2 \equiv\sum_{i=1}^N (y_i-m_i)^2/\sigma^2_i,
\end{equation}
where $m_i \equiv m(t_i,\meanparams)$, with respect to $\meanparams$. Where does this come from?

@@

\subsubsection{Link to Gaussian Process Regression}\label{sec:gp-lsq-link}
\dan{I think we should explicitly re-write this as $\mathcal{L}(\meanparams,\,\hyperparams)$ and indicate that $K$ and $m$ depend on these parameters. We should really highlight that this is a drop-in replacement for the likelihood defined in the previous section.} \suz{Is this ok?}
Let us now re-write the likelihood in matrix form:
\begin{equation}
\label{eq:gp_like}
  \mathcal{L}(\meanparams,\hyperparams) = 
  \frac{1}{\sqrt{|2\,\pi\,\mathbf{K}|}} \exp \left(-(\mathbf{y}-\mathbf{m})^{\mathrm{T}} \mathbf{K}^{-1}
  (\mathbf{y}-\mathbf{m}) \right),
\end{equation}
where, as before, the mean vector $\mathbf{m}$ has elements $m_i=m(t_i,\meanparams)$ and $\mathbf{K}$ is a purely diagonal ($N$, $N$) matrix with elements $K_{ij} = \delta_{ij} \sigma_i^2$ ($\delta_{ij}$ being the discrete Kronecker delta function). This is, of course, the \emph{covariance matrix} of the model.

Now, instead of assuming that the covariance matrix takes this very specific form, let us allow a more flexible covariance model:
\begin{equation}
  K_{ij} = k(t_i,t_j,\hyperparams) + \delta_{ij} \sigma_i^2,
\end{equation}
where $k$ is a \emph{covariance function} or \emph{kernel}, controlled by parameters $\hyperparams$. The result is a GP, and its likelihood is still given by Equation~(\ref{eq:gp_like}). Depending on the choice of kernel and parameters, the covariance matrix can now have non-zero off-diagonal elements, allowing us to explicitly model correlated noise or stochastic signals in the data.
Note that the likelihood depends not only on the argument of the exponential, but also on the determinant of the covariance matrix, which therefore needs to be evaluated explicitly. This term acts as a built-in Occam's razor, automatically penalising more complex models.

@@

\subsubsection{Properties of the  predictive distribution}
A closer look at Equations~(\ref{eq:pred}) reveals some important properties.
First, a GP is a \textit{linear predictor}. The predictive mean for a specific time $t_\star$ can be written as a linear combination of the observations:
$f_\star = \mathbf{w}^{\mathrm{T}} \, \mathbf{y}$, where $\mathbf{w} = \mathbf{K}^{-1} \, \mathbf{k}_\star$ and $\mathbf{k}_\star=k(\mathbf{t},t_\star,\hyperparams)$. It can also be written as
a linear combination of covariance functions centred on the training points:
$f_\star = \mathbf{\alpha}^\mathrm{T} \,  \mathbf{k}_\star$ where $\mathbf{\alpha} = \mathbf{K}^{-1} \, \mathbf{y}$.
These linearity properties are very important to understand the behaviour of GPs. They shed light on the relationship between GPR and standard linear models with very large numbers of free parameters \citep[see e.g.][for a more detailed discussion]{2021PASP..133i3001H},
and lead to a number of powerful extensions that are beyond  scope of this review.

@@

In the previous section, we presented an overview of GP methods, and the key mathematical details. In this section, we will dive deeper into some of the practical decisions that arise when using GPs.
The two core elements of a GP model are the mean function $m(t;\,\meanparams)$, and the ``kernel'' or ``covariance'' function $k(t_i,\,t_j;\,\hyperparams)$.
Unlike how they are typically presented in the machine learning literature, GP models in astrophysics will often---but not always---include non-trivial mean functions.
For example, in the example from \autoref{sec:transit}, the mean function $m(t;\,\meanparams)$ is a physical transit model that is a function of the orbital parameters of the system, and includes a realistic limb-darkening model \citep{2002ApJ...580L.171M}.
However, in this review we won't discuss the mean function in detail, focusing instead primarily on the kernel function, since that is unique to GP modeling.
All this being said, in our experience, new users of GP models will often focus and worry more than necessary about the choice of kernel function for their problem.
As with any probabilistic modeling problem, there are several well-defined workflows for motivating, selecting, and validating your choice of kernel.
In this section we walk through this process in detail.

@@

The kernel function can be any positive semi-definite scalar function, but some will be more useful than others.
% A commonly (over-)used kernel function is the ``exponential-squared'' or ``radial basis function'' kernel
% \begin{equation}
%   k(t_i,\,t_j;\,\hyperparams) = \sigma^2\,\exp\left[-\frac{(t_i - t_j)^2}{2\,\tau^2}\right]
% \end{equation}
% where the ``hyperparameters'' $\hyperparams$ are $\sigma$ and $\tau$.
% In this case, the hyperparameters specify the typical amplitude $\sigma$ and timescale $\tau$ of the variance of the process.
% SOMEFIGURE shows how changing these parameters affect our GP model.
% There exist many other commonly used kernel functions, some of which are listed in SOMETABLE.
% This table is far from comprehensive, and importantly, the sum or product of two valid kernel functions is itself also a valid kernel, so it is possible (and common) to build extremely flexible and expressive kernel functions specific to your use case.
Given this large decision space, you may be wondering how to choose the right kernel for your specific problem.
If you're very lucky, you may be able to motivate your model using physics.
This can be approached from two directions.
On one hand, some commonly used kernel function have a specific physical interpretation---for example, the solution to a stochastic ordinary differential equation---that can be used to motivate their use.
On the other hand, if your physical model is specified by its \emph{power spectrum}, this can recast as a GP model with a specific kernel function. \suz{Say something about the number of times a function is differentiable? That can be a useful consideration. But perhaps already mentioned later.}

@@

Some popular kernels are listed in \autoref{tab:kernels}, and some other choices are discussed in Chapter 4 of \citet{gpml}.
In any practical application, these kernel functions are not used on their own.
Instead, more expressive models are designed by combining these models.
In particular, any valid kernel functions can be added or multiplied to generate new valid functions.
For example, the squared exponential kernel function is generally defined as
\begin{equation}
  k(\dt;\,\hyperparams) = \amplitude^2\,\exp\left(-\frac{\dt^2}{2\,\lengthscale^2}\right)
\end{equation}
with $\hyperparams = \{\amplitude,\,\lengthscale\}$, which, in our notation, is actually the product of a constant kernel and a squared exponential kernel, as defined in \autoref{tab:kernels}.
Another example that is commonly used in the astrophysics literature \citep{2012MNRAS.419.3147A,2014MNRAS.443.2517H} is the following ``quasi-periodic'' kernel
\begin{equation}
  k(\dt;\,\hyperparams) = \amplitude^2\,\exp\left(-\frac{\dt^2}{2\,{\lengthscale_1}^2} -\gamma\,\sin^2\left[\frac{\pi\,\dt}{\lengthscale_2}\right] \right)
\end{equation}
with $\hyperparams = \{\amplitude,\,\lengthscale_1,\,\lengthscale_2,\,\gamma\}$, which has a period of $\lengthscale_2$, and a decoherence timescale of $\lengthscale_1$.

@@

\suz{Mention kernels with break points to detect sudden changes, as described in review by \citet{2012RSPTA.37110550R} for example?}

@@

\subsection{Stars and exoplanets}
\label{sec:stars_planets}

@@

\subsubsection{Transit searches}
\label{sec:transit_det}

@@

\subsubsection{Transit and eclipse modelling}
\label{sec:transit_fit}
